Thermodynamic response functions in a cell fluid model

### 2024-09-01 main_morse_response_functions.tex
change terminology from chain rule to cyclic relation
explicitly write that there are two eos - P(T,mu) and P(T,n)

	\section{The equation of state}
	\subsection{Pressure as a function of temperature and chemical potential}
	The equation of state obtained in~\cite[see Eq.(42)]{KozlovskiiDobush2020} reads
	\begin{equation}
		\label{eq:eosMT}
		Pv\beta = E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4.
	\end{equation}
	The quantities in the left-hand side of the equation are $P$, the pressure; $\beta = (k_{\rm B} T)^{-1}$, the inverse temperature; $k_{\rm B}$, the Boltzmann constant; $T$, the temperature; $v$, cell volume. The quantities in the right-hand side are, in general, functions of the temperature $T$ and the chemical potential $\mu$. Let us present their expressions explicitly.

@@

	Let us introduce the reduced pressure
	\begin{equation}
		\label{def:reduced_pres}
		P^* \equiv \frac{P v}{\varepsilon}.
	\end{equation}
	Considering the equation of state~\eqref{eq:eosMT}, $P^*$ can be written as
	\begin{equation}
		\label{eq:reduced_pres}
		P^* = \frac{1 + \tau}{\varepsilon \beta_c} Pv\beta.
	\end{equation}
	This equation can be easily represented graphically. Figure~\ref{fig1} illustrates the relationship between the reduced pressure $P^*$ and the effective chemical potential $M$ for various values of the reduced temperature $\tau$, at and above the critical temperature. At the critical point $M=0, \tau=0$, $P^*_c = 1.606$.

@@

	This dependency can also be expressed explicitly. To achieve this, we combine the equations~\eqref{eq:density} and~\eqref{eq:ro_M} to express the effective chemical potential $M$ as a function of density and temperature:
	\begin{equation}\label{eq:M_nT}
		\bar M = \frac{\rho_{n}}{\beta W(0)} - (\bar n - n_g),
	\end{equation}
	where
	\begin{align} \label{eq:ro_nT}
		 & \rho_{n} = - 2 \left(\frac{g_3^2 - 2g_2 g_4}{g_4^2} \right)^{\frac{1}{2}} \cos \left( \frac{\alpha_n}{3} + \frac{\pi}{3} \right), \\
		 & \alpha_n = \arccos \left( \frac{3}{2} \left( - \frac{g_4^4}{\left( 2 g_2 g_4 - g_3^2\right)^3}\right)^{\frac{1}{2}} (n_g - \bar n)\right). \nonumber
	\end{align}

	At $T>T_c$ the equation of state of a cell fluid model in terms of the density and the temperature has the following form
	\begin{equation}\label{eq:eosNT}
	Pv\beta = E_\eta (\bar n,T) + \bar M \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4,
	\end{equation}
 The function $E_\nu (\bar n, T)$ in \eqref{eq:ro_nT} is the function $E_\mu (M,T)$ \eqref{eq:E_mu} rewritten in terms of density and temperature taking into account the expression \eqref{eq:M_nT}.

	Figure~1 shows the isotherms for the pressure $P$ following from the equation of state ~\eqref{eq:eosMT} in terms of (chemical potential, temperature)  and \eqref{eq:eosNT} in terms of (density, temperature).
	\begin{figure}[h!]
		%\centering 
		\includegraphics[width=0.446\textwidth]{f1a1.pdf} 
		\includegraphics[width=0.45\textwidth]{f1b1.pdf} 
		\vskip-3mm\caption{Isotherms of the reduced pressure $P^*$ as a function of the density $\bar n$ (on the left-hand side) and a function of the effective chemical potential $M$ (on the right-hand side) at $T \geq T_c$ represented by black lines. Thick grey lines on both figures correspond to isotherms of pressure at $T < T_c$ based on the results taken from~\cite{KozlovskiiDobush2020}. 
		}\label{fig1}
	\end{figure}

	\section{Thermodynamic response functions (coefficients)}

@@

	Let us introduce the reduced isothermal compressibility
	\begin{equation}
		K^*_T \equiv \frac{\varepsilon K_T}{v}.
	\end{equation}
	The quantity $K^*_T$ is dimensionless and is of order unity, except at the critical point itself, where it is divergent. Figure~\ref{fig2a} illustrates the dependency of the compressibility $K^*_T$ on the density $\bar{n}$ for various values of temperature above the critical one.

@@

	\subsection{Thermal pressure coefficient}
	The thermal pressure coefficient is defined by
	\begin{equation}
		\label{def:therm_pres_coef}
		\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}.
	\end{equation}
	We rewrite $\beta_V$ into a form that is suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray}
		\beta_V & = & \left(\frac{\partial P}{\partial T}\right)_{V, N}
		\nonumber\\
		& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
	\end{eqnarray}
	Applying the chain rule
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
		\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we obtain
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		& = & 
		- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
		\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
		& = & - \left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}.		
	\end{eqnarray*}
	Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
	\begin{equation}
		\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
		- \left(\frac{\partial P}{\partial \mu}\right)_T 
		\left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
	(NB. Compare this equation with Eq.(17) from~\cite{StrokerMeier2021})

@@

	\subsection{Thermal expansion coefficient}
	The thermal expansion coefficient is defined by
	\begin{equation}
		\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
	\end{equation}
	We rewrite $\alpha_P$ into a form suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray*}
		\alpha_P & = & \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
		\\
		& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{P, N}
		\\
		& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{\mu} 
		 - \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
		 \left(\frac{\partial \mu}{\partial T}\right)_{P, N}.
	\end{eqnarray*}
	Applying the chain rule
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{P, N}
		\left(\frac{\partial T}{\partial P}\right)_{\mu, N}
		\left(\frac{\partial P}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we get
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{P, N} & = & 
		- \left(\frac{\partial P}{\partial T}\right)_{\mu}
		\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}.
	\end{eqnarray*}
	Substituting the result into the expression for $\alpha_P$, we arrive at the final formula for the thermal expansion coefficient
	\begin{equation}
		\alpha_P = -\frac{1}{\rho}\left(\frac{\partial \rho}{\partial T}\right)_{\mu}
		+ \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
		\left(\frac{\partial P}{\partial T}\right)_{\mu}
		\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is readily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
	(NB. Compare this equation with Eqs.(57)-(58) from~\cite{StrokerMeier2021})

>>>>

	\section{\label{sec:eos}The equation of state}
	\subsection{Pressure as a function of temperature and chemical potential}
	The equation of state obtained in~\cite[see Eq.(42)]{KozlovskiiDobush2020} reads
	\begin{equation}
		\label{eq:eosMT}
		Pv\beta = E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4.
	\end{equation}
	The quantities in the left-hand side of the equation are $P$, the pressure; $\beta = (k_{\rm B} T)^{-1}$, the inverse temperature; $k_{\rm B}$, the Boltzmann constant; $T$, the temperature; $v$, cell volume. The quantities in the right-hand side are, in general, functions of the temperature $T$ and the chemical potential $\mu$. Let us present their expressions explicitly.

@@

	Let us introduce the reduced pressure
	\begin{equation}
		\label{def:reduced_pres}
		P^* \equiv \frac{P v}{\varepsilon}.
	\end{equation}
	Considering the equation of state~\eqref{eq:eosMT}, $P^*$ is explicitly written as
	\begin{equation}
		\label{eq:eosPTM_reduced}
		P^* = \frac{1 + \tau}{\varepsilon \beta_c} \left( E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4 \right).
	\end{equation}
	This equation can be easily represented graphically. Figure~\ref{fig1} illustrates the relationship between the reduced pressure $P^*$ and the effective chemical potential $M$ for various values of the reduced temperature $\tau$, at and above the critical temperature. At the critical point $M=0, \tau=0$, $P^*_c = 1.606$.

@@

	This dependency can also be expressed explicitly. To achieve this, we combine the equations~\eqref{eq:density} and~\eqref{eq:ro_M} to express the effective chemical potential $M$ as a function of density and temperature:
	\begin{equation}\label{eq:M_nT}
		\bar M = \frac{\rho_{n}}{\beta W(0)} - (\bar n - n_g),
	\end{equation}
	where
	\begin{align} \label{eq:ro_nT}
		 & \rho_{n} = - 2 \left(\frac{g_3^2 - 2g_2 g_4}{g_4^2} \right)^{\frac{1}{2}} \cos \left( \frac{\alpha_n}{3} + \frac{\pi}{3} \right), \\
		 & \alpha_n = \arccos \left( \frac{3}{2} \left( - \frac{g_4^4}{\left( 2 g_2 g_4 - g_3^2\right)^3}\right)^{\frac{1}{2}} (n_g - \bar n)\right). \nonumber
	\end{align}
	{\color{Red} {\bf{NOTE 2: }} Roma: In my opinion, under {\it{arccos}} we should have 3, not $\frac{3}{2}$}

	At $T>T_c$ the equation of state of a cell fluid model in terms of the density and the temperature has the following form
	\begin{equation}\label{eq:eosNT}
	Pv\beta = E_\eta (\bar n,T) + \bar M \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4,
	\end{equation}
 The function $E_\nu (\bar n, T)$ in \eqref{eq:ro_nT} is the function $E_\mu (M,T)$ \eqref{eq:E_mu} rewritten in terms of density and temperature taking into account the expression \eqref{eq:M_nT}. For the reduced pressure we write
 \begin{equation}
 	\label{eq:eosPTn_reduced}
 	P^* = \frac{1 + \tau}{\varepsilon \beta_c} \left(E_\eta (\bar n,T) + \bar M \rho_{n} + \frac{\tilde D(0)}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4\right).
 \end{equation}

	Figure~1 shows the isotherms for the pressure $P^*$ versus density $\bar{n}$ (left-hand side) and versus the effective chemical potential $M$ (right-hand side).
	\begin{figure}[h!]
		%\centering 
		\includegraphics[width=0.446\textwidth]{f1a1.pdf} 
		\includegraphics[width=0.45\textwidth]{f1b1.pdf} 
		\vskip-3mm\caption{Isotherms of the reduced pressure $P^*$ as a function of the density $\bar n$ (on the left-hand side) and a function of the effective chemical potential $M$ (on the right-hand side) at $T \geq T_c$ represented by black lines. Thick grey lines on both figures correspond to isotherms of pressure at $T < T_c$ based on the results taken from~\cite{KozlovskiiDobush2020}. 
		}\label{fig1}
	\end{figure}

	Thus, in this Section, we presented two forms of the equation of state. The first one expresses the pressure as a function of temperature and chemical potential, $P = P(T, \mu)$, and is represented by Eqs.~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}. The second one expresses the pressure as a function of temperature and density, $P = P(T, \bar{n})$, and is represented by Eqs.~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}. These equations, together with the explicit dependency of density on temperature and chemical potential, $\bar{n} = \bar{n}(T, \mu)$, see Eq.~\eqref{eq:density}, are the basis for the calculation of the thermodynamic response functions in the next Section~\ref{sec:respons_functions}

	\section{\label{sec:respons_functions}Thermodynamic response functions}

@@

	Let us introduce the reduced isothermal compressibility
	\begin{equation}
		K^*_T \equiv \frac{\varepsilon K_T}{v}.
	\end{equation}
	The quantity $K^*_T$ is dimensionless and is of order unity, except at the critical point itself, where it is divergent. It is expressed in terms of the reduced quantities $P^*$ and $\bar{n}$ as follows
	\begin{eqnarray}
		K^*_T & = & \frac{\varepsilon}{\bar{n}^2} \left(\frac{\partial \bar{n}}{\partial \mu}\right)_T
		\nonumber\\
		& = & \frac{\varepsilon^2}{\bar{n}^2} \left(\frac{\partial^2 P^*}{\partial \mu^2}\right)_T.
	\end{eqnarray}
	Now, either Eq.~\eqref{eq:eosPTM_reduced} or Eq.~\eqref{eq:density} can be used to explicitly calculate $K^*_T$, with the result expressed as a function of temperature and chemical potential.

	Figure~\ref{fig2a} illustrates the dependency of the compressibility $K^*_T$ on the density $\bar{n}$ for various values of temperature above the critical one. The dependency of $K^*_T$ on the effective chemical potential $M$ is displayed in Fig~\ref{fig2b}.

@@

	\subsection{Thermal pressure coefficient}
	The thermal pressure coefficient is defined by
	\begin{equation}
		\label{def:therm_pres_coef}
		\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}.
	\end{equation}
	We rewrite $\beta_V$ into a form that is suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray}
		\beta_V & = & \left(\frac{\partial P}{\partial T}\right)_{V, N}
		\nonumber\\
		& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
	\end{eqnarray}
	Applying the cyclic relation
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
		\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we obtain
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
		& = & 
		- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
		\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
		& = & - \left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}.		
	\end{eqnarray*}
	Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
	\begin{equation}
		\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
		- \left(\frac{\partial P}{\partial \mu}\right)_T 
		\left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
		\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
	(NB. Compare this equation with Eq.(17) from~\cite{StrokerMeier2021})

@@

	\subsection{Thermal expansion coefficient}
	The thermal expansion coefficient is defined by
	\begin{equation}
		\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
	\end{equation}
	We rewrite $\alpha_P$ into a form suitable for the equation of state~\eqref{eq:eosMT}
	\begin{eqnarray*}
		\alpha_P & = & \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
		\\
		& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{P, N}
		\\
		& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{\mu} 
		 - \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
		 \left(\frac{\partial \mu}{\partial T}\right)_{P, N}.
	\end{eqnarray*}
	Applying the cyclic relation
	\begin{equation*}
		\left(\frac{\partial \mu}{\partial T}\right)_{P, N}
		\left(\frac{\partial T}{\partial P}\right)_{\mu, N}
		\left(\frac{\partial P}{\partial \mu}\right)_{T, N}
		= -1
	\end{equation*}
	we get
	\begin{eqnarray*}
		\left(\frac{\partial \mu}{\partial T}\right)_{P, N} & = & 
		- \left(\frac{\partial P}{\partial T}\right)_{\mu}
		\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}.
	\end{eqnarray*}
	Substituting the result into the expression for $\alpha_P$, we arrive at the final formula for the thermal expansion coefficient
	\begin{equation}
		\alpha_P = -\frac{1}{\rho}\left(\frac{\partial \rho}{\partial T}\right)_{\mu}
		+ \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
		\left(\frac{\partial P}{\partial T}\right)_{\mu}
		\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}
	\end{equation}
	which is readily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
	(NB. Compare this equation with Eqs.(57)-(58) from~\cite{StrokerMeier2021})
